---
title: "Spirals Dataset"
output:
  bookdown::pdf_document2:
    citation_package: default
    latex_engine: xelatex
    toc: false
    includes:
        in_header: "preamble.tex"
---

# Introduction
```{r, warning = FALSE, message = FALSE, echo = FALSE}
# import necessary libraries
library(mlbench)
library(ggplot2)
library(dplyr)
library(scatterplot3d)
library(viridis)
theme_set(theme_bw())
```
The spirals dataset is a synthetic dataset from `mlbench`. The spirals problem's inputs are points on two entangled spirals. If sd is greater than zero, Gaussian noise is introduced to each data point.

To generate the dataset, we need to define `n` - number of patterns to create, `cycles` - the number of cycles each spiral makes, and `sd` - standard deviation of data points around the spirals. Below is an example of the spirals dataset with one cycle each (`cycles = 1`) and no noise (`sd = 0`):

```{r, message = FALSE, warning = FALSE}
# load the dataset from mlbench
spirals_1 <- mlbench.spirals(n = 300, cycles = 1, sd = 0) %>% as_tibble()
print(spirals_1)
```
Basically, the spirals dataset will always have 2 features `x.1` and `x.2` indicating the 2-d coordinates of the data points and one target `classes` indicating which class each data point belongs to. Because there is no added noise, the data points from each spiral will align perfectly on the curve.
```{r, message = FALSE, warning = FALSE}
viridis_2 <- viridis(2, end = 0.9)
ggplot(data = spirals_1, mapping = aes(color = classes, x = x.1, y = x.2)) +
        geom_point() +
        scale_color_manual(values = viridis_2)
```

We can also easily generate another spirals dataset with a more complex parameters: 2 cycles and `sd = 0.1`:

```{r, message = FALSE, warning = FALSE}
spirals_2 <- mlbench.spirals(n = 300, cycles = 2, sd = 0.1) %>% as_tibble()
ggplot(data = spirals_2, mapping = aes(color = classes, x = x.1, y = x.2)) +
        geom_point() +
        scale_color_manual(values = viridis_2)
```
