---
title: "Lecture File Enumerator"
author: SLDS Lecture Service
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(data.table)
library(dplyr)
library(fs)
library(reactable)
library(ggplot2)

knitr::opts_chunk$set(warning = FALSE, message = FALSE)

lecture_dir <- here::here()
```

## File Enumeration

This document enumerates all files in this lecture repository with their sizes and extensions.  


- Certain hidden directories like `.git` and `.Rproj.user` are ignored automatically, but false-positives for "irrelevant" files are possible.  
- If you see too many LaTeX compilation leftovers, run `make clean` before rendering this document.
- File extensions are normalized to lower case to avoid spurious duplicates
- Dotfiles (files starting with `.` are given counted as `(dotfile)` extensions)
- Files without extension (like `Makefile`) are counted as `(no extension)`

```{r enumerate-files}
enumerate_files <- function(lecture_path) {
  files <- fs::dir_ls(lecture_path, all = TRUE, recurse = TRUE, type = "file")

  # Make paths relative to lecture directory
  files_rel <- fs::path_rel(files, start = lecture_path)

  # Filter out .git
  files_rel <- fs::path_filter(files_rel, glob = "**.git/*", invert = TRUE)
  files_rel <- fs::path_filter(
    files_rel,
    glob = "**.Rproj.user*",
    invert = TRUE
  )
  files_rel <- fs::path_filter(
    files_rel,
    glob = "file_counts_files**",
    invert = TRUE,
  )

  extension <- tolower(fs::path_ext(files_rel))
  is_dotfile <- stringr::str_detect(
    fs::path_file(files_rel),
    pattern = "^\\.[\\w\\d]*"
  )
  extension[is_dotfile] <- "(dotfile)"
  extension[extension == ""] <- "(no extension)"

  data.table(
    file = fs::path_file(files_rel),
    parent_dir = fs::path_dir(files_rel),
    extension = extension,
    size = fs::file_size(files),
    size_kb = as.numeric(fs::file_size(files)) / 1024,
    size_mb = as.numeric(fs::file_size(files)) / 1024 / 1024
  )
}

file_dt <- enumerate_files(lecture_dir)
file_dt[, depth := lengths(fs::path_split(parent_dir))]
```

### Summary Statistics

```{r summary}
total_files <- nrow(file_dt)
total_size_mb <- sum(file_dt$size_mb)
```

- Total files: **`r total_files`**
- Total size: **`r round(total_size_mb, 2)`MB**

### File Count by Extension

```{r extension-counts}
ext_counts <- file_dt[,
  .(
    count = .N,
    total_size_mb = sum(size_mb)
  ),
  by = extension
][order(-count)]

ext_counts[1:20] |>
  reactable(
    columns = list(
      extension = colDef(name = "Extension"),
      count = colDef(name = "Files", format = colFormat(separators = TRUE)),
      total_size_mb = colDef(name = "Size (MB)", format = colFormat(digits = 2))
    ),
    sortable = TRUE,
    filterable = TRUE,
    defaultPageSize = 20
  )
```

### File Count by Directory

```{r dir-counts}
dir_counts <- file_dt[,
  .(
    file_count = .N,
    total_size_mb = sum(size_mb),
    depth = first(depth)
  ),
  by = parent_dir
][order(-file_count)]

dir_counts |>
  reactable(
    columns = list(
      parent_dir = colDef(name = "Directory", minWidth = 300),
      file_count = colDef(
        name = "Files",
        format = colFormat(separators = TRUE)
      ),
      total_size_mb = colDef(
        name = "Size (MB)",
        format = colFormat(digits = 2)
      ),
      depth = colDef(name = "Depth")
    ),
    sortable = TRUE,
    filterable = TRUE,
    defaultPageSize = 20
  )
```

### Largest Files

```{r largest-files}
file_dt[order(-size_mb)][1:30] |>
  mutate(full_path = file.path(parent_dir, file)) |>
  select(depth, full_path, size_mb, extension) |>
  reactable(
    columns = list(
      depth = colDef(name = "Depth"),
      full_path = colDef(name = "File Path", minWidth = 400),
      size_mb = colDef(name = "Size (MB)", format = colFormat(digits = 3)),
      extension = colDef(name = "Extension")
    ),
    sortable = TRUE,
    filterable = TRUE,
    defaultPageSize = 30
  )
```

### Extensions by Directory Depth

```{r depth-extension}
depth_ext <- file_dt[,
  .(
    count = .N,
    total_size_mb = sum(size_mb)
  ),
  by = .(depth, extension)
][order(depth, -count)]

depth_ext |>
  filter(count > 5) |>
  reactable(
    columns = list(
      depth = colDef(name = "Depth"),
      extension = colDef(name = "Extension"),
      count = colDef(name = "Files", format = colFormat(separators = TRUE)),
      total_size_mb = colDef(name = "Size (MB)", format = colFormat(digits = 2))
    ),
    sortable = TRUE,
    filterable = TRUE,
    defaultPageSize = 20
  )
```

## Distributions

```{r size-distribution}
ggplot(file_dt[size_kb > 0], aes(x = size_kb)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  scale_x_log10(labels = scales::comma) +
  labs(
    title = "File Size Distribution",
    x = "File Size (KB, log scale)",
    y = "Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title.position = "plot")
```


```{r extension-chart}
#| fig-height: 7
file_dt[, .N, by = extension] |>
  ggplot(aes(y = reorder(extension, N), x = N)) +
  geom_col() +
  labs(
    title = "File Extension Distribution",
    x = "No. of files",
    y = "Extension"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title.position = "plot")
```
