---
title: "Exercise 2 -- Regression"
subtitle: "[Introduction to Machine Learning](https://slds-lmu.github.io/i2ml/)"
notebook-view:
  - notebook: ex_regression_R.ipynb
    title: "Exercise sheet for R"
    url: "https://github.com/slds-lmu/lecture_i2ml/blob/master/exercises/supervised-regression/ex_regression_R.ipynb"
  - notebook: ex_regression_py.ipynb
    title: "Exercise sheet for Python"
    url: "https://github.com/slds-lmu/lecture_i2ml/blob/master/exercises/supervised-regression/ex_regression_py.ipynb"
  - notebook: sol_regression_R.ipynb
    title: "Solutions for R"
    url: "https://github.com/slds-lmu/lecture_i2ml/blob/master/exercises/supervised-regression/sol_regression_R.ipynb"
  - notebook: sol_regression_py.ipynb
    title: "Solutions for Python"
    url: "https://github.com/slds-lmu/lecture_i2ml/blob/master/exercises/supervised-regression/sol_regression_py.ipynb"
---

::: {.content-hidden when-format="pdf"}
::: {.hidden}
{{< include ../_quarto/latex-math.qmd >}}
:::
{{< include ../_quarto/solution-buttons.qmd >}}
:::

<details>
<summary>*Hint: Useful libraries*</summary>

::: {.panel-tabset}

### R
{{< embed ex_regression_R.ipynb#import echo=true >}}
### Python
{{< embed ex_regression_py.ipynb#import echo=true >}}
:::
</details>

## Exercise 1: Loss functions for regression tasks

::: {.callout-note title="Learning goals" icon=false}
1. Assess how outliers affect models for different loss functions
2. Derive impact of a loss function from visual representation
:::

In this exercise, we will examine loss functions for regression tasks
somewhat more in depth.

```{r, echo=FALSE, fig.height=3, fig.width=5}
set.seed(1L)
x <- runif(20L, min = 0L, max = 10L)
y <- 0.2 + 3 * x
y <- y + rnorm(length(x), sd = 0.8)

ggplot2::ggplot(data.frame(x = x, y = y), ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_point() +
  ggplot2::theme_bw() +
  ggplot2::annotate("point", x = 10L, y = 1L, color = "orange", size = 2)
```

a) Consider the above linear regression task. How will the model parameters be affected by adding the new outlier point (orange) if you use $L1$ loss and $L2$ loss, respectively, in the empirical risk? (You do not need to actually compute the parameter values.)

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details>
<summary>**Solution**</summary>

$L2$ loss penalizes vertical distances to the regression line *quadratically*, while $L1$ only considers the *absolute* distance. As the outlier point lies pretty far from the remaining training data, it will have a large loss with $L2$, and the regression line will pivot to the bottom right to minimize the resulting empirical risk. A model trained with $L1$ loss is less susceptible to the outlier and will adjust only slightly to the new data.

```{r, echo=FALSE, fig.height=3, fig.width=5, message=FALSE}
set.seed(1L)
x <- runif(20L, min = 0L, max = 10L)
y <- 0.2 + 3 * x
y <- y + rnorm(length(x), sd = 0.8)
x <- c(x, 10L)
y <- c(y, 1L)
dt <- data.frame(x = x, y = y)

ggplot2::ggplot(dt, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_point() +
  ggplot2::theme_bw() +
  ggplot2::geom_smooth(
    formula = y ~ x,
    method = "lm",
    se = FALSE,
    ggplot2::aes(col = "L2 loss"),
    linewidth = 0.7
  ) +
  ggplot2::geom_quantile(quantiles = 0.5, ggplot2::aes(col = "L1 loss")) +
  ggplot2::scale_color_manual(
    "", values = c("L2 loss" = "blue", "L1 loss" = "red")
  ) +
  ggplot2::annotate("point", x = 10L, y = 1L, color = "orange", size = 2)
```

</details>
:::
:::

b) The second plot visualizes another loss function popular in regression tasks, the so-called *Huber loss* (depending on $\epsilon > 0$; here: $\epsilon = 5$). Describe how the Huber loss deals with residuals as compared to $L1$ and $L2$ loss. Can you guess its definition?

::: {.indent-1}
```{r, echo=FALSE, fig.height=3, fig.width=5}
huber_loss <- function(res, delta = 0.5) {
  if (abs(res) <= delta) {
    0.5 * (res^2)
  } else {
    delta * abs(res) - 0.5 * (delta^2)
  }
}

x <- seq(-10L, 10L, length.out = 1000L)
y <- sapply(x, huber_loss, delta = 5L)

ggplot2::ggplot(data.frame(x = x, y = y), ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_line() +
  ggplot2::theme_bw() +
  ggplot2::labs(x = "y - f(x)", y = "L(y, f(x)")
```
:::

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details>
<summary>**Solution**</summary>
The Huber loss combines the respective advantages of $L1$ and $L2$ loss: it is smooth and (once) differentiable like $L2$ but does not punish larger residuals as severely, leading to more robustness. It is simply a (weighted) piecewise combination of both losses, where $\epsilon$ marks where $L2$ transits to $L1$ loss. The exact definition is:

$$
\Lxy = \begin{cases}
\frac{1}{2}(y - \fx)^2  & \text{ if } \lone \le \epsilon \\
\epsilon \lone-\frac{1}{2}\epsilon^2 \quad & \text{ otherwise }
\end{cases}, \quad \epsilon > 0
$$

In the plot we can see how the parabolic shape of the loss around 0 evolves into an absolute-value function at $\lone > \epsilon = 5$.
</details>
:::
:::

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details>
<summary>**Illustration: $L1$, $L2$, and Huber loss side by side**</summary>

To tie a) and b) together, we revisit the regression task from a) and add the model fitted with Huber loss. Like $L1$, the Huber model barely reacts to the outlier -- its line stays close to the robust $L1$ fit (here the two practically coincide), whereas $L2$ is visibly dragged towards the outlier in the bottom right. Huber thus inherits the robustness of $L1$ while -- being differentiable everywhere (recall its smooth shape in b) -- remaining easy to optimize.

```{r, echo=FALSE, fig.height=3, fig.width=5, message=FALSE, warning=FALSE}
set.seed(1L)
x <- runif(20L, min = 0L, max = 10L)
y <- 0.2 + 3 * x
y <- y + rnorm(length(x), sd = 0.8)
x <- c(x, 10L)
y <- c(y, 1L)
dt <- data.frame(x = x, y = y)

ggplot2::ggplot(dt, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_point() +
  ggplot2::theme_bw() +
  ggplot2::geom_smooth(
    formula = y ~ x,
    method = "lm",
    se = FALSE,
    ggplot2::aes(col = "L2 loss"),
    linewidth = 0.7
  ) +
  ggplot2::geom_quantile(quantiles = 0.5, ggplot2::aes(col = "L1 loss")) +
  ggplot2::geom_smooth(
    formula = y ~ x,
    method = MASS::rlm,
    se = FALSE,
    ggplot2::aes(col = "Huber loss"),
    linewidth = 0.7,
    linetype = "dashed"
  ) +
  ggplot2::scale_color_manual(
    "",
    values = c("L2 loss" = "blue", "L1 loss" = "red", "Huber loss" = "darkgreen")
  ) +
  ggplot2::annotate("point", x = 10L, y = 1L, color = "orange", size = 2)
```

</details>
:::
:::

## Exercise 2: HRO in coding frameworks

::: {.callout-note title="Learning goals" icon=false}
1. Translate lecture concepts to code
2. Evaluate a model using a custom loss function
:::

Throughout the lecture, we will frequently use the `R` package `mlr3`, resp. the `Python` package `sklearn`, and its descendants, providing an integrated ecosystem for all common machine learning tasks. Let's recap the HRO principle and see how it is reflected in either `mlr3` or `sklearn`. An overview of the most important objects and their usage, illustrated with numerous examples, can be found at [the `mlr3` book](https://mlr3book.mlr-org.com/) and [the `scikit` documentation](https://scikit-learn.org/stable/index.html).

a) How are the key concepts (i.e., hypothesis space, risk and optimization) you learned about in the lecture videos implemented?

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details>
<summary>**Solution**</summary>

::: {.panel-tabset}
### R
{{< embed sol_regression_R.ipynb#hro-objects echo=true >}}
### Python
{{< embed sol_regression_py.ipynb#hro-objects echo=true >}}
:::
</details>
:::
:::

b) Have a look at `mlr3::tsk("iris")` / `sklearn.datasets.load_iris`. What attributes does this object store?

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details>
<summary>**Solution**</summary>

::: {.panel-tabset}
### R
{{< embed sol_regression_R.ipynb#hro-task echo=true >}}
### Python
{{< embed sol_regression_py.ipynb#hro-task echo=true >}}
:::
</details>
:::
:::

c) Instantiate a regression tree learner (`lrn("regr.rpart")` / `DecisionTreeRegressor`). What are the different settings for this learner?

   <details>
   <summary>*Hint*</summary>
   ::: {.panel-tabset}

   ### R
   `mlr3::mlr_learners$keys()` shows all available learners.

   ### Python
   Use `get_params()` to see all available settings.
   :::
   </details>

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details>
<summary>**Solution**</summary>

::: {.panel-tabset}
### R
{{< embed sol_regression_R.ipynb#hro-learner echo=true >}}

```{=latex}
\needspace{20\baselineskip}
```

{{< embed sol_regression_R.ipynb#hro-learner-params echo=true >}}
### Python
{{< embed sol_regression_py.ipynb#hro-learner echo=true >}}
:::
</details>
:::
:::

d) In Exercise 1, you encountered the *Huber loss*. Unlike $L1$ and $L2$ loss, it is not shipped as a ready-made loss/measure in either `mlr3` or `sklearn`. An implementation (following the definition from Exercise 1) is provided below. Use it to compute the empirical Huber risk (the mean Huber loss over all residuals) of the linear model from a). How does it compare to the MSE, and what does the difference tell you about how the two measures treat large residuals? In `sklearn`, you can wrap the loss into a custom scorer via `make_scorer`.

::: {.indent-1}
::: {.panel-tabset}
### R
{{< embed sol_regression_R.ipynb#hro-huber echo=true eval=false >}}
### Python
{{< embed sol_regression_py.ipynb#hro-huber echo=true eval=false >}}
:::
:::

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details>
<summary>**Solution**</summary>

::: {.panel-tabset}
### R
{{< embed sol_regression_R.ipynb#hro-huber-risk echo=true >}}
### Python
{{< embed sol_regression_py.ipynb#hro-huber-risk echo=true >}}
:::

Both measures score the *same* model, but weight errors differently: the MSE ($\approx 15$) squares the residuals and is therefore dominated by the largest ones, whereas the Huber risk ($\approx 2.6$) grows only linearly beyond $\epsilon$ and keeps their influence in check. The comparison thus turns the robustness we saw visually in Exercise 1 into a concrete number -- large residuals count for much less under Huber than under $L2$.

The R and Python values differ slightly because the synthetic data in a) is drawn with each language's own random number generator: the same seed does not produce the same sample in R and Python, so the two models are fitted to (slightly) different data.
</details>
:::
:::

::: {.content-hidden when-profile="b"}

## Exercise 3: Polynomial regression
::: {.callout-tip icon=false title="The whole exercise is only for lecture group A!"}
:::

::: {.callout-note title="Learning goals" icon=false}
1. Express HRO components for polynomial regression
2. Derive gradient update for optimization
3. Analyze and discuss learner flexibility
:::

Assume the following (noisy) data-generating process from which we have observed 50 realizations: $$y = -3 + 5 \cdot \sin(0.4 \pi x) + \epsilon$$ with $\epsilon \, \sim \mathcal{N}(0, 1)$.

```{r, echo=FALSE, fig.height=3, fig.width=5}
set.seed(123L)
x <- seq(-3, 3, length.out = 50)
y <- -3 + 5 * sin(0.4 * pi * x) + rnorm(50, sd = 1)

ggplot2::ggplot(data.frame(x = x, y = y), ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_point() +
  ggplot2::theme_bw()
```

a) We decide to model the data with a cubic polynomial (including intercept term). State the corresponding hypothesis space.

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details>
<summary>**Solution**</summary>
Cubic means degree 3, so our hypothesis space will look as follows:

$$
\Hspace = \{ \fxt = \theta_0 + \theta_1 x + \theta_2 x^2 + \theta_3 x^3
~|~ (\theta_0, \theta_1, \theta_2, \theta_3)^\top \in \R^4 \}
$$

</details>
:::
:::

b) State the empirical risk w.r.t. $\thetav$ for a member of the hypothesis space. Use $L2$ loss and be as explicit as possible.

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details>
<summary>**Solution**</summary>
The empirical risk is:

$$
\risket = \sum_{i = 1}^{50} \left(\yi - \left[ \theta_0 + \theta_1 x^{(i)} +
\theta_2 \left( x^{(i)} \right)^2 + \theta_3 \left( x^{(i)} \right)^3 \right]
\right)^2
$$

</details>
:::
:::

c) We can minimize this risk using gradient descent. Derive the gradient of the empirical risk w.r.t $\thetav$.

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details>
<summary>**Solution**</summary>
We can find the gradient just as we did for an intermediate result when we derived the least-squares estimator:

\begin{align*}
  \nabla_{\thetav} \risket &=
  \pd{}{\thetav} \left \| \yv - \Xmat \thetav \right \|_2^2 \\
  &= \pd{}{\thetav} \left( \left(\yv - \Xmat \thetav\right)^\top
  \left(\yv - \Xmat \thetav \right) \right) \\
  &= - 2 \yv^\top \Xmat + 2 \thetav^\top \Xmat^\top \Xmat\\
  &= 2 \cdot \left(-\yv^\top \Xmat + \thetav^\top \Xmat^\top \Xmat \right)
\end{align*}

</details>
:::
:::

d) Using the result for the gradient, explain how to update the current parameter $\thetat$ in a step of gradient descent.

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details>
<summary>**Solution**</summary>
Recall that the idea of gradient descent (\textit{descent}!) is to traverse the risk surface in the direction of the \textit{negative} gradient as we are in search for the minimum. Therefore, we will update our current parameter set $\thetat$ with the negative gradient of the current empirical risk w.r.t. $\thetav$, scaled by learning rate (or step size) $\alpha$:

$$
\thetatn = \thetat - \alpha \cdot \nabla_{\thetav} \riske(\thetat).
$$

What actually happens here: we update each component of our current parameter vector $\thetat$ in the \textit{direction} of the negative gradient, i.e., following the steepest downward slope, and also by an \textit{amount} that depends on the value of the gradient.

In order to see what that means it is helpful to recall that the gradient $\nabla_{\thetav} \risket$ tells us about the effect (infinitesimally small) changes in $\thetav$ have on $\risket$. Therefore, gradient updates focus on  influential components, and we proceed more quickly along the important dimensions.
</details>
:::
:::

e) You will not be able to fit the data perfectly with a cubic polynomial. Describe the advantages and disadvantages that a more flexible model class would have. Would you opt for a more flexible learner?

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details>
<summary>**Solution**</summary>
We see that, for example, the first model in exercise b) fits the data fairly well but not perfectly. Choosing a more flexible function (a polynomial of higher degree or a function from an entirely different, more complex, model class) might be advantageous:

* We would be able to trace the observations more closely if our function were less smooth, and thus reduce empirical risk. On the other hand, flexibility also has drawbacks:

* Flexible model classes often have more parameters, making training harder.

* We might run into a phenomenon called *overfitting*. Recall that our ultimate goal is to make predictions on *new* observations. However, fitting every quirk of the training observations -- possibly caused by imprecise measurement or other factors of randomness/error -- will not generalize so well to new data.

In the end, we need to balance model fit and generalization. We will discuss the choice of hypotheses quite a lot since it is one of the most crucial design decisions in machine learning.

</details>
:::
:::

:::

::: {.content-hidden when-profile="a"}

## Exercise 4: Fitting a linear model by hand
::: {.callout-tip icon=false title="The whole exercise is only for lecture group B!"}
:::

::: {.callout-note title="Learning goals" icon=false}
1. Compute the least-squares estimator on a small example by hand
2. Interpret linear model coefficients
3. Predict with a fitted linear model
:::

In the lecture, you saw that a linear model with $L2$ loss has a closed-form solution -- the *ordinary-least-squares (OLS)* estimator
$$
\thetavh = \olsest,
$$
where $\Xmat$ is the design matrix (the feature columns padded with a leading column of ones for the intercept) and $\yv$ the vector of observed targets. Consider the following small dataset with a single feature $x$ and target $y$:

| $i$ | $x^{(i)}$ | $y^{(i)}$ |
|:---:|:---:|:---:|
| 1 | $-3$ | $1$ |
| 2 | $-1$ | $1$ |
| 3 | $1$  | $3$ |
| 4 | $3$  | $7$ |

We model it with a simple linear regression $f(x) = \theta_0 + \theta_1 x$.

a) Set up the design matrix $\Xmat$ and compute the OLS estimator $\thetavh = (\hat{\theta}_0, \hat{\theta}_1)^\top$ by hand.

   <details>
   <summary>*Hint*</summary>

   The feature is centered ($\sum_i x^{(i)} = 0$), which makes $\Xmat^\top \Xmat$ diagonal and hence trivial to invert.

   </details>

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details>
<summary>**Solution**</summary>

With the intercept column, the design matrix and target vector are
$$
\Xmat = \begin{pmatrix} 1 & -3 \\ 1 & -1 \\ 1 & 1 \\ 1 & 3 \end{pmatrix}, \qquad
\yv = \begin{pmatrix} 1 \\ 1 \\ 3 \\ 7 \end{pmatrix}.
$$

We assemble the two ingredients of the OLS estimator:
$$
\Xmat^\top \Xmat =
\begin{pmatrix} n & \sum_i x^{(i)} \\ \sum_i x^{(i)} & \sum_i \left(x^{(i)}\right)^2 \end{pmatrix}
= \begin{pmatrix} 4 & 0 \\ 0 & 20 \end{pmatrix},
\qquad
\Xmat^\top \yv =
\begin{pmatrix} \sum_i y^{(i)} \\ \sum_i x^{(i)} y^{(i)} \end{pmatrix}
= \begin{pmatrix} 12 \\ 20 \end{pmatrix}.
$$

Because the feature is centered ($\sum_i x^{(i)} = 0$), $\Xmat^\top \Xmat$ is diagonal, so its inverse simply inverts the diagonal entries:
$$
\left(\Xmat^\top \Xmat\right)^{-1} = \begin{pmatrix} \tfrac{1}{4} & 0 \\ 0 & \tfrac{1}{20} \end{pmatrix}.
$$

Hence
$$
\thetavh = \left(\Xmat^\top \Xmat\right)^{-1} \Xmat^\top \yv
= \begin{pmatrix} \tfrac{1}{4} & 0 \\ 0 & \tfrac{1}{20} \end{pmatrix}
\begin{pmatrix} 12 \\ 20 \end{pmatrix}
= \begin{pmatrix} 3 \\ 1 \end{pmatrix},
$$
so the fitted model is $f(x) = 3 + x$.

</details>
:::
:::

b) Interpret the two fitted coefficients $\hat{\theta}_0$ and $\hat{\theta}_1$.

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details>
<summary>**Solution**</summary>

- Intercept $\hat{\theta}_0 = 3$: the predicted target when $x = 0$. As the feature is centered, $x = 0$ is the mean feature value, so the intercept here coincides with the mean target $\bar{y} = \tfrac{1}{4}(1 + 1 + 3 + 7) = 3$.
- Slope $\hat{\theta}_1 = 1$: the *marginal effect* of the feature -- increasing $x$ by one unit increases the predicted target by one unit (ceteris paribus).

</details>
:::
:::

c) Use the fitted model to predict the target for a new observation with $x = 2$.

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details>
<summary>**Solution**</summary>

$$
\hat{y} = \hat{\theta}_0 + \hat{\theta}_1 \cdot 2 = 3 + 1 \cdot 2 = 5.
$$

</details>
:::
:::

:::

## Exercise 5: Predicting `abalone`
::: {.callout-tip icon=false title="Bonus exercise -- will not be discussed in the tutorial session"}
This optional exercise is meant for both lecture groups.
:::

::: {.callout-note title="Learning goals" icon=false}
1. Implement a regression model
2. Interpret linear model coefficients
3. Analyze basic regression fit
:::

We want to predict the age of an abalone using its longest shell measurement and
its weight. The `abalone` data can be found here: [https://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.data](https://archive.ics.uci.edu/ml/machine-learning-databases/abalone/abalone.data).

Prepare the data as follows:

::: {.panel-tabset}
### R
{{< embed sol_regression_R.ipynb#abalone-data echo=true eval=false >}}

### Python
{{< embed sol_regression_py.ipynb#abalone-data echo=true eval=false >}}
:::

a) Plot `LongestShell` and `WholeWeight` on the $x$- and $y$-axis, respectively, and color points according to `Rings`.

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details>
<summary>**Solution**</summary>
::: {.panel-tabset}
### R
{{< embed sol_regression_R.ipynb#abalone-plotr echo=true >}}
### Python
{{< embed sol_regression_py.ipynb#abalone-plot echo=true >}}
:::

We see that weight scales exponentially with shell length and that larger/heavier animals tend to have more rings.
</details>
:::
:::

b) Using `mlr3`/`sklearn`, fit a linear regression model to the data.

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details>
<summary>**Solution**</summary>
::: {.panel-tabset}
### R
<!-- {{< embed sol_regression_R.ipynb#abalone-task echo=true >}} -->
{{< embed sol_regression_R.ipynb#abalone-predict echo=true >}}
### Python
{{< embed sol_regression_py.ipynb#abalone-predict echo=true >}}
:::
</details>
:::
:::

c) A key advantage of the linear model is that its parameters are directly interpretable. Read off the fitted coefficients and interpret the effect of `longest_shell` on the predicted number of rings.

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details>
<summary>**Solution**</summary>
The fitted linear model reads
$$
\hat{f}(\xv) = 3.43 + 10.58 \cdot x_{\text{shell}} + 1.16 \cdot x_{\text{weight}}.
$$

The coefficient of `longest_shell` ($\approx 10.58$) is the *marginal effect* of this feature: increasing `longest_shell` by one unit -- while holding `whole_weight` fixed -- increases the predicted number of rings (i.e., the estimated age) by about $10.58$. Both coefficients are positive, matching our observation from a) that larger and heavier animals tend to have more rings. Note that `longest_shell` only ranges from roughly $0.1$ to $0.8$, so a full-unit change is large relative to the observed data.

::: {.panel-tabset}
### R
{{< embed sol_regression_R.ipynb#abalone-coef echo=true eval=false >}}
### Python
{{< embed sol_regression_py.ipynb#abalone-coef echo=true eval=false >}}
:::
</details>
:::
:::

d) Compare the fitted and observed targets visually.

   <details>
   <summary>*R Hint*</summary>
   ::: {.panel-tabset}
   ### R
   Use `$autoplot()` from `mlr3viz`.
   :::
   </details>

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details>
<summary>**Solution**</summary>
::: {.panel-tabset}
### R
{{< embed sol_regression_R.ipynb#abalone-vizr echo=true >}}
### Python
{{< embed sol_regression_py.ipynb#abalone-viz echo=true >}}
:::

We see a scatterplot of prediction vs true values, where the small bars along the axes (a so-called rugplot) indicate the number of observations that fall into this area. As we might have suspected from the first plot, the underlying relationship is not exactly linear (ideally, all points and the resulting line should lie on the diagonal). With a linear model we tend to underestimate the response.

</details>
:::
:::

e) Assess the model's training loss in terms of MAE.

   <details>
   <summary>*Hint*</summary>
   ::: {.panel-tabset}
   ### R
   Call `$score()`, which accepts
   different `mlr_measures`, on the prediction object.

   ### Python
   Call `from sklearn.metrics import mean_absolute_error`.
   :::
   </details>

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details>
<summary>**Solution**</summary>
::: {.panel-tabset}
### R
{{< embed sol_regression_R.ipynb#abalone-eval echo=true >}}
### Python
{{< embed sol_regression_py.ipynb#abalone-eval echo=true >}}
:::
</details>
:::
:::
