---
title: "Content"
output: rmarkdown::html_vignette
toc:
  depth: 2
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{"Content"}
---



 

## Learning Compendia

For each topic, we provide videos, online quizzes, and small programming 
exercises in R. They are combined in a *learning compendium*. 
Links to the individual videos and slides can be found below.


- [ML Basics](https://compstat-lmu.shinyapps.io/tutorial-ml-basics/)
- [Supervised Regression](https://compstat-lmu.shinyapps.io/tutorial-supervised-regression/)
- [Supervised Classification](https://compstat-lmu.shinyapps.io/tutorial-supervised-classification/)
- [Evaluation](https://compstat-lmu.shinyapps.io/tutorial-evaluation/)
- [Trees](https://compstat-lmu.shinyapps.io/tutorial-trees/)
- [Random Forests](https://compstat-lmu.shinyapps.io/tutorial-forests/)
- [Tuning](https://compstat-lmu.shinyapps.io/tutorial-tuning/)


<!-- - [](https://compstat-lmu.shinyapps.io/tutorial-/) -->

---



## Videos and Slides

In this section you can find the links to all videos individually as well as
the slides used in the videos.

```{r, echo=FALSE, eval=TRUE, message=FALSE}
source("get_slide_data.R")
library("knitr")

kable(subset(sldat, select = c("Lesson", "Description", "Topic", "YouTube", "PDF")), 
    row.names = FALSE)
```

## Machine Learning framework mlr3 

We use the mlr3 framework for many applied ML exercises. 
Please read the section under [References](../articles/references.html) regarding the background material. 

## Code Demos

You can run the code snippets in the demos on your local machine.
If you want to render the Rmd files to PDF, you need the accompanying [style files](https://github.com/compstat-lmu/lecture_i2ml/tree/master/style).

```{r, echo=FALSE, eval=TRUE, message=FALSE}
source("get_demo_data.R")

kable(subset(cddat, select = c("Topic", "Demo", "Rmd", "PDF")), row.names = FALSE)
```

Some of these code demos reference each other, so if you want to run them yourself and experiment with the code (...you definitely should!), make sure you download them all to the same folder.  


## Exercises

```{r, echo=FALSE, eval=TRUE, message=FALSE}
source("get_exercise_data.R")

kable(exdat, row.names = FALSE)
```

## Cheat sheets

```{r, echo=FALSE, eval=TRUE, message=FALSE}
source("get_cheatsheet_data.R")

kable(cdat, row.names = FALSE)
```

More cheat sheets are in the making and will be added soon.
