\exercise

```{r, echo=FALSE, cache=FALSE}
if (! exists("solution")) {
  solution = FALSE
}
```

The following data set contains information about the *shoe size* and *body height* of 3 persons:

```{r}
shoes = data.frame(shoe.size = c(46, 40, 44), body.height = c(180, 172, 175))
shoes
```

  a) Normalize the data.\
  **Hint:** have a look at `?scale`
  a) Compute the manhattan and the euclidean distance between the 1. and 3. observation by hand
      - for the normalized data 
      - for the original `shoes` data
  a) Solve the previous exercise with R. \
  **Hint:** have a look at `?dist`
  a) Compute the coordinates of the centroid of the original `shoes` data using all observations.

```{r, results="asis", echo=FALSE, eval=solution}
cat("__Solution__:")
```

```{r, message=FALSE, echo = solution, eval = solution}
# a)
##########
shoes.norm = scale(shoes)
shoes.norm

# alternative: if you don't know the scale function, write your own.
scale_this = function(x) {
  (x - mean(x)) / sd(x)
}
apply(shoes, 2, scale_this)

# b)
##########
diff.norm = shoes.norm[1, ] - shoes.norm[3, ]
diff.norm 
# manhattan for normalized data:
sum(abs(diff.norm))
# euclidean for normalized data:
sqrt(sum((diff.norm)^2))

diff = shoes[1, ] - shoes[3, ]
diff
# manhattan for original data:
sum(abs(diff))
# euclidean for normalized data:
sqrt(sum((diff)^2))

# c)
##########
# manhattan for normalized data:
as.numeric(dist(shoes.norm[c(1,3),], method = "manhattan"))
# euclidean for normalized data:
as.numeric(dist(shoes.norm[c(1,3),], method = "euclidean"))

# manhattan for original data:
as.numeric(dist(shoes[c(1,3),], method = "manhattan"))
# euclidean for normalized data:
as.numeric(dist(shoes[c(1,3),], method = "euclidean"))

# d)
##########
shoes
# coordinates of the centroid are
centroid = c(mean(shoes$shoe.size), mean(shoes$body.height))
centroid
```

```{r, echo = FALSE, eval = solution}
plot(rbind(shoes, shoes[1,]), type = "b")
points(x = centroid[1], y = centroid[2], pch = 19)
```

