---
output: pdf_document
params:
    set_title: "mlr3 Usecase - Ames Housing Dataset"
---

---
title: `r params$set_title`
department: "Department of Statistics -- LMU Munich"
author:
  - Working Group "Computational Statistics" -- Bernd Bischl et al.
pdf_document:
    fig_caption: yes
    highlight: pygments
    keep_tex: yes
    number_sections: yes
    toc: yes
urlcolor: darkgray
header-includes:
  - \usepackage[english]{babel}
  - \usepackage[utf8]{inputenc}
  - \usepackage{dsfont}
  - \usepackage{verbatim}
  - \usepackage{amsmath}
  - \usepackage{amsfonts}
  - \usepackage{csquotes}
  - \usepackage{multirow}
  - \usepackage{longtable}
  - \usepackage{enumerate}
  - \usepackage[absolute,overlay]{textpos}
  - \usepackage{psfrag}
  - \usepackage{algorithm}
  - \usepackage{algpseudocode}
  - \usepackage{eqnarray}
  - \usepackage{arydshln}
  - \usepackage{tabularx}
  - \usepackage{placeins}
  - \usepackage{tikz}
  - \usepackage{setspace}
  - \usepackage{colortbl}
  - \usepackage{mathtools}
  - \usepackage{wrapfig}
  - \usepackage{subfig}
---

\centering

\raggedright


```{r, child = "../style/setup.Rmd", include = FALSE}

```


This usecase is intended to use mlr3 to perform a benchmark analysis on the _Ames Dataset_.

The main objectives of this study are as follows:

- To apply data preprocessing and preparation techniques in order to obtain clean data
- To build machine learning models able to predict house price based on house features
- To analyze and compare models performance in order to choose the best model

## Accessing the dataset

The dataset is available on Kaggle https://www.kaggle.com/prevek18/ames-housing-dataset.
Kaggle is a platform which provides data science competitions and datasets which can be used to get familiar with typical machine learning methods.

## Importing the data

```{r}
housing = read.csv("data/AmesHousing.csv")
```

Load mlr3

```{r}
library(mlr3)
library(mlr3learners)

task = TaskRegr$new(id = "ames_housing", backend = as_data_backend(housing), target = "SalePrice")
task

lrn = mlr_learners$get("regr.ranger")
lrn$train(task)
```
