---
output: pdf_document
params:
  set_title: "Code Demo for Resampling"
---
  
```{r resampling-preamble, child = "../style/preamble_code_demos.Rmd", include = FALSE, purl = FALSE}

```

```{r resamplings-setup, child = "../style/setup.Rmd", include = FALSE, purl = FALSE}

```  

# Code demo for resampling

## Self made Cross Validation

We want to assess the performance of our model, i.e., try to estimate its generalization error. 
Why is it a good idea to use cross-validation (CV)?

Let's write our very own CV function for a $k$-NN learner to experiment with....
```{r resampling-knn_cv}
library(mlr3)
library(mlr3learners)
library(mlbench)

set.seed(13)
spiral <- as.data.frame(mlbench.spirals(n = 500, sd = 0.1))

# Cross validation for kNN
# inputs:
#   data a data set to use
#   target name of the column in <data>to classify
#   folds number of CV folds
#   k neighborhood size for kNN
# returns vector of test set errors for the different folds
knn_cv <- function(data, target, folds, k) {
  cv_errors <- as.numeric(folds)

  indices <- c(
    sample(x = seq(1, nrow(data), by = 1), size = nrow(data), replace = FALSE),
    rep(NA, (folds - nrow(data) %% folds) %% folds)
  )

  # index matrix for folds
  index_mat <- matrix(data = indices, byrow = FALSE, nrow = folds)

  for (i in 1:folds) {
    # data
    test_data <- data[na.omit(index_mat[i, ]), ]
    train_data <- data[-na.omit(index_mat[i, ]), ]
    task <- TaskClassif$new(
      id = "spirals_train",
      backend = train_data, target = target
    )
    # model
    learner <- lrn("classif.kknn", k = k)
    # train on training set
    learner$train(task = task)
    # evaluate on test data
    cv_errors[i] <- learner$predict_newdata(test_data)$score()
  }

  cv_errors
}

result <- knn_cv(data = spiral, target = "classes", folds = 11, k = 4)
result
mean(result)
```

```{r resampling-knn_cv_plot}
p <- ggplot(data = as.data.frame(result), aes(y = result)) +
  geom_point(x = 0) +
  ggtitle(label = "CV error") +
  xlab("") + ylab("test error") + xlim(c(0, 0))
p
```

So what happens if we increase the number of folds?

```{r resampling-knn_cv_folds}
# run CV with 2, 3, 4, 5, 10, 15, ...., 60 folds and record the test set errors.
cv_results <- lapply(
  X = c(2, 3, 4, 5 * (1:12)),
  FUN = function(folds) {
    data.frame(
      folds = as.character(folds),
      cv_errors = knn_cv(
        data = spiral, target = "classes",
        folds = folds, k = 4
      )
    )
  }
)
cv_data <- do.call(rbind, cv_results)


cv_plot <-
  ggplot(cv_data, aes(x = folds, y = cv_errors)) + geom_boxplot() +
  ggtitle(label = "CV with different no. of folds") +
  xlab("number of folds") + ylab("test error")
cv_plot
```

The more we increase the number of folds, the larger each training set becomes. Hence the *pessimistic bias* for the estimated model performance becomes smaller.  
But since the test sets also become smaller, the *variance* of the resulting performance estimate increases.  
In addition, with a higher number of folds, the computation time increases. 
(Think about that: by how much? is the  increase linear in the number of folds? why or why not?) 

Can we get better results with a smaller amount of computation?
Let's see what happens if we do repeated CV and collect only their means:

```{r resampling-knn_x_cv}
# do reps = 10 repetitions each of 5, 10, 15, 20-fold CV
rep_cv_results <- lapply(
  X = c(5, 10, 15, 20),
  FUN = function(folds, reps = 10) {
    mean_cv_errors <- replicate(
      reps,
      mean(knn_cv(
        data = spiral, target = "classes",
        folds = folds, k = 4
      ))
    )
    data.frame(
      folds = as.character(folds),
      mean_cv_errors = mean_cv_errors
    )
  }
)

rep_cv_data <- do.call(rbind, rep_cv_results)

ggplot(rep_cv_data, aes(x = folds, y = mean_cv_errors)) +
  geom_boxplot() +
  xlab("10 times repeated x-CV") + ylab("test error") +
  ggtitle(label = "Repeated CV (10 reps) with different no. of folds") +
  ylim(range(cv_data$cv_errors))
```

We see that our estimation results stabilize.

## `mlr3`'s CV implementation

```{r resampling-mlr}
set.seed(1337)
task <- TaskClassif$new(
  id = "spirals_task",
  backend = spiral, target = "classes"
)
rdesc_cv <- rsmp("repeated_cv", folds = 10, repeats = 10)

mlr_cv <- resample(
  resampling = rdesc_cv, learner = lrn("classif.kknn", k = 4),
  task = task
)

mlr_cv$score()[, c("iteration", "classif.ce")]
mlr_cv$aggregate()
```

```{r resampling-mlr_plot}
library(ggplot2)

ggplot(data = mlr_cv$score()[, "classif.ce"], aes(y = classif.ce, x = 1)) +
  geom_boxplot() +
  ggtitle(label = "Repeated CV (10-10) with mlr") +
  xlab("") + ylab("test error") + xlim(c(0, 2))
```
