# 5th Notebook for Machine Learning Intro Lecture

# 1. introductory kaggle challenge

We will compete in our first [kaggle challenge](https://www.kaggle.com/c/titanic) on the prediction of titanic survivors. 

### Preprocessing and Data check

```{r}
### Data preprocess

# load and check the data
allTrain = read.csv(file = "data/train_titanic.csv")
str(allTrain)
# no target column "survived" on test dataset
allTest = read.csv(file = "data/test_titanic.csv")

# can we use all features? 
# Nope: delete those with too many levels as this would inflate the model 
# also kill the ID

train = allTrain[, -c(which(colnames(allTrain) == "Cabin"), 
    which(colnames(allTrain) == "Name"), 
    which(colnames(allTrain) == "Ticket"), 
    which(colnames(allTrain) == "PassengerId"))]

test = allTest[, -c(which(colnames(allTest) == "Cabin"), 
    which(colnames(allTest) == "Name"), 
    which(colnames(allTest) == "Ticket"), 
    which(colnames(allTest) == "PassengerId"))]
```

### Build a first simple model with mlr and check the performance via CV

```{r, warning=FALSE}
### model corner
library(mlr)

# choose specific model and parameters 
task = makeClassifTask(data = train, target = "Survived")

learner = makeLearner(cl = "classif.rpart")
# check choosable parameters and set accordingly
getLearnerParamSet(learner)
# check available settings here: https://www.rdocumentation.org/packages/rpart/versions/4.1-12/topics/rpart.control
learner = makeLearner(cl = "classif.rpart", 
    par.vals = list(minsplit = 10, cp = 0.05))

# make sure to assign mlr package to train
mod = mlr::train(learner = learner, task = task)

### performance estimate via CV
cv = makeResampleDesc(method = "CV", iters = 10)
# use mlr::listMeasures() to get list of possible measures
# important: always check on which measure they evaluate you!
mlr::crossval(learner = learner, task = task, iters = 10, measures = list(mmce, acc))
```

Store and submit your predictions

```{r}
# predict for submission
pred = predict(mod, newdata = test)
submission = pred$data

submission$PassengerId = allTest$PassengerId

colnames(submission) = c("Survived", "PassengerId")

write.csv(submission, file = "data/submissionTitanic_1.csv", row.names = FALSE)
```

### Tune the Hyperparameters of the algorithm

```{r}
### Tune the model 
# we chose two numeric parameters above and now search for optimal values
# check available parameters
set.seed(1337)
getLearnerParamSet(learner)
# make parameter set
paramSet = makeParamSet(
    makeDiscreteParam("minsplit", values = c(1, 3, 5, 7, 10, 15, 20, 30, 40, 45, 50, 60, 70, 100)), 
    makeNumericParam("cp", lower = 0.0001, upper = 0.1)
)
# choose random search - why not grid search?
ctrl = makeTuneControlRandom(maxit = 100)
resDesc = makeResampleDesc("CV", iters = 10L, predict = "both")
tunes = mlr::tuneParams(learner = learner, task = task, resampling = resDesc,
    par.set = paramSet, control = ctrl, measures = list(mmce, acc))
```

Visualize the random search over both parameters:

```{r}
visHyper = generateHyperParsEffectData(tunes)
plt = plotHyperParsEffect(visHyper, x = "minsplit", y = "cp", z = "acc.test.mean")
plt
# tuning result
tunes
```

Store and submit those results to kaggle

```{r}
# use those param settings for the CART
learner = makeLearner(cl = "classif.rpart", 
    par.vals = list(minsplit = 10, cp = 0.0923))# inspect the learner
# learner
mod = mlr::train(learner = learner, task = task)

# predict for submission
pred = predict(mod, newdata = test)
submission = pred$data

submission$PassengerId = allTest$PassengerId

colnames(submission) = c("Survived", "PassengerId")

write.csv(submission, file = "data/submissionTitanic_2.csv", row.names = FALSE)
```

#### Check variable importances

```{r}
library(ggplot2)
varimp = getFeatureImportance(mod)
var = as.data.frame(t(varimp$res))
var$names = names(varimp$res)
p = ggplot(data = var, aes(x = names, y = V1)) + geom_bar(stat = "identity") +
    ggtitle(label = "Variable Importances")
p
```

### Feature engineering

Can we further condense the information from the multi-level factors and use it for our model?

We take a closer look at the names of the guests. 

```{r, message=FALSE, warning=FALSE}
### feature engineering
library(dplyr)

# indicator for train or test set
allTrain$train = 1
allTest$train = 0
allTest$Survived = NA

# compute once for all data and split again for training with ID
allData = rbind(allTrain, allTest)
engData = allData

head(allData$Name)
```

We can see, that there is information on the title of the people in their names. We use that information as a new feature!

```{r}
# use regular expressions via strplit to extract the title of the people
# temporary storage
temp = sapply(strsplit(as.character(allData$Name), split = ","), function(x) x[2])
title = strsplit(temp, split = " ")
engData$title = sapply(title, function(x) x[2])
# unfortunately still too many titles with too few observations
table(engData$title)
```

Btw.: we found the Captain:

```{r}
# btw.: we found the captain:
allData[which(engData$title == "Capt."), "Name"]
```

condense those with obs < 5 to class "other"

```{r}
freqs = as.data.frame(table(engData$title))
otherTitles = freqs[which(freqs$Freq < 5), "Var1"]
engData[which(engData$title %in% otherTitles), "title"] = "other"
engData$title = as.factor(engData$title)
# looks better now
table(engData$title)
```

### Build updated model

```{r}
### model corner 2 with engineered feature
library(mlr)
library(dplyr)

train = engData %>% 
    filter(train == 1) %>%
    select(-c(PassengerId, Name, Ticket, train, Cabin))

test = engData %>% 
    filter(train == 0) %>%
    select(-c(PassengerId, Name, Ticket, train, Cabin, Survived))

# choose specific model and parameters 
task = makeClassifTask(data = train, target = "Survived")

learner = makeLearner(cl = "classif.rpart")
# we chose two numeric parameters above and now search for optimal values
paramSet = makeParamSet(
    makeDiscreteParam("minsplit", values = c(1, 3, 5, 7, 10, 15, 20, 30, 40, 45, 50, 60, 70, 100)), 
    makeNumericParam("cp", lower = 0.0001, upper = 0.1)
)
# choose random search - why not grid search?
ctrl = makeTuneControlRandom(maxit = 100)
resDesc = makeResampleDesc("CV", iters = 10L, predict = "both")
tunes = mlr::tuneParams(learner = learner, task = task, resampling = resDesc,
    par.set = paramSet, control = ctrl, measures = list(mmce, acc))
```

Check tuning result

```{r}
tunes
```

Write and store the submission

```{r}
# use those param settings for the CART
learner = makeLearner(cl = "classif.rpart", 
    par.vals = list(minsplit = 7, cp = 0.00601))

mod = mlr::train(learner = learner, task = task)

# predict for submission
pred = predict(mod, newdata = test)
submission = pred$data

submission$PassengerId = engData %>% 
    filter(train == 0) %>%
    select(PassengerId)

colnames(submission) = c("Survived", "PassengerId")

write.csv(submission, file = "data/submissionTitanic_3.csv", row.names = FALSE)
```

#### Check Variable Importances

```{r}
library(ggplot2)
varimp = getFeatureImportance(mod)
var = as.data.frame(t(varimp$res))
var$names = names(varimp$res)
p = ggplot(data = var, aes(x = names, y = V1)) + geom_bar(stat = "identity") +
    ggtitle(label = "Variable Importances")
p
```

What can we see? How could we critisize that result? Is there a way to detect the problem?

```{r}
```