## Predict Probabilities with mlr
```{r, include=FALSE}
library(mlr)
```

1. Set `predict.type = "prob"` in `makeLearner`:
    ```{r}
    lrn.lda = makeLearner("classif.lda", predict.type = "prob")
    ```

2. Train and predict model as explained:
    ```{r}
    mod.prob = train(lrn.lda, iris.task)
    preds = predict(mod.prob, iris.task)
    ```

3. Get probabilities by using `getPredictionProbabilities`:
    ```{r, size="tiny"}
    head(getPredictionProbabilities(preds))
    ```

## Predict Probabilities with mlr


Not every learner has a pre defined method to predict probabilities:
```{r, eval=FALSE, size="tiny"}
lrn.knn = makeLearner("classif.knn", predict.type = "prob")
## Error in setPredictType.Learner(learner = wl, predict.type = predict.type) :
##   Trying to predict probs, but classif.knn does not support that!
```

| Method              |        Learner       | `predict.type` |
| ------------------- | -------------------- |:--------------:|
| k-NN                | `classif.knn`        | No             |
| k-NN                | `classif.kknn`       | Yes            |
| QDA                 | `classif.qda`        | Yes            |
| LDA                 | `classif.lda`        | Yes            |
| Naive Bayes         | `classif.naiveBayes` | Yes            |
| Logistic Regression | `classif.logreg`     | Yes            |
