---
title: "Exercise 9 -- Random Forests"
subtitle: "[Introduction to Machine Learning](https://slds-lmu.github.io/i2ml/)"
notebook-view:
  - notebook: ex_forests_R.ipynb
    title: "Exercise sheet for R"
    url: "https://github.com/slds-lmu/lecture_i2ml/blob/master/exercises/forests/ex_forests_R.ipynb"
  - notebook: ex_forests_py.ipynb
    title: "Exercise sheet for Python"
    url: "https://github.com/slds-lmu/lecture_i2ml/blob/master/exercises/forests/ex_forests_py.ipynb"
  - notebook: sol_forests_R.ipynb
    title: "Solutions for R"
    url: "https://github.com/slds-lmu/lecture_i2ml/blob/master/exercises/forests/sol_forests_R.ipynb"
  - notebook: sol_forests_py.ipynb
    title: "Solutions for Python"
    url: "https://github.com/slds-lmu/lecture_i2ml/blob/master/exercises/forests/sol_forests_py.ipynb"
---

::: {.content-hidden when-format="pdf"}
::: {.hidden}
{{< include ../_quarto/latex-math.qmd >}}
:::
{{< include ../_quarto/solution-buttons.qmd >}}
:::

<details> 
<summary>*Hint: Useful libraries*</summary>

::: {.panel-tabset}
### R
{{< embed ex_forests_R.ipynb#import echo=true >}}
### Python
{{< embed ex_forests_py.ipynb#import echo=true >}}
:::
</details>

## Exercise 1: Bagging
::: {.callout-tip icon=false title="This exercise is only for lecture group A"}
:::

::: {.callout-note title="Learning goals" icon=false}
1. Understand benefit of bagging from a mathematical perspective
2. Solve "show that..."-type exercises
3. Handle expectations over random variables
:::

In this exercise, we briefly revisit why bagging is a useful technique to stabilize predictions.

a) For a fixed observation $(\xv, y)$, show that the expected quadratic loss over individual
base learner predictions $\blx$ is larger than or equal to the quadratic loss of the prediction $\fM$ of a size-$M$ ensemble.
You can consider all hyperparameters of the base learners and the ensemble fixed.
   <details> 
   <summary>*Hint*</summary>
 
   Use the law of total expectation ("Verschiebungssatz der Varianz": $\var(Z) =
   \E(Z^2) - (\E(Z))^2 ~ \Longleftrightarrow ~ \E(Z^2) = \var(Z) + (\E(Z))^2$,
   where $\var(Z) \geq 0$ by definition.) stating $\E(Z^2) \geq (\E(Z))^2$ for a random variable $Z$.
 
   </details>

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details> 
<summary>**Solution**</summary>

- Start with the LHS of the inequality: the *expected quadratic loss over individual
  base learner predictions* $\rightsquigarrow$ $\E \left( \left(y - \blx \right)^2 \right)$
- Get clear about RHS: the *quadratic loss of the prediction of a size-$M$
  ensemble* $\rightsquigarrow$ $\left( y - \left(\fM \right) \right)^2$
- Before we get busy moving from LHS to RHS, let's think about the
  expectation for a minute.

::: {.callout-note title="Distribution of the loss"}
**Expected value** of a RV: (possibly infinite) sum over all values the RV could take,
weighted
by the probability of observing that value.

What even is the RV here? Since the base learner and ensemble
structure is fixed for given data, the only stochastic part is the bootstrap sample
in the $m$-th tree given training data from the data-generating process.
We could write this as $\E_{\Dtrain^{[m]} ~|~ \Dtrain
\sim \Pxy}$.
In this exercise we'll omit the subscript for ease of notation.
:::

- Back to our proof, which will make use of the LOTV.
- By the LOTV, we know that $\E(Z^2) \geq (\E(Z))^2$ for some RV $Z$.
- Applying that to our LHS, we obtain
 $$
\E \left( \left(y - \blx \right)^2\right) \geq \left(\E \left( y - \blx \right) \right)^2 .
$$
- Expected values of larger terms can often be simplified so the expectation is only
  over the actually stochastic parts (using *linearity* of expectation), yielding:
$$
\E \left( \left(y - \blx \right)^2\right) \geq \left( y - \E \left(\blx \right) \right)^2 .
$$
- The last missing step is to show that $\E \left(\blx \right) = \fM$. To compute the expectation for this discrete random variable (we have a finite ensemble), we sum over all possible realizations, weighted by their probability of occurence.
This can be further simplified given that all of the $M$ bootstrap samples were drawn with equal probability:

$$
\E \left(\blx \right) = \sum_{m = 1}^M \blx p \left(\Dtrain^{[m]} \right)
= \frac{1}{M} \sum_{m = 1}^M \blx,
$$

which is precisely the ensemble prediction $\fM$.

Putting everything together, we get 

$$
\E \left( \left(y - \blx \right)^2\right) \geq \left( y - \left(\fM \right) \right)^2,
$$

showing that the expected quadratic loss over individual base learner predictions is at least as large as the loss of the ensemble prediction.

</details> 
:::
:::

## Exercise 2: Classifying spam

::: {.callout-note title="Learning goals" icon=false}
1) Apply RF to data for prediction, OOB error estimation & feature importance computation
2) Understand how 63% probability for observations to end up in a tree comes about
:::

a) ::: {.callout-tip icon=false title="Only for lecture group B"}
   Take a look at the `spam` dataset and shortly describe what kind of classification problem this is. \textcolor{teal}{[only for lecture group B]}
   :::

   <details> 
   <summary>*Hint*</summary>
   ::: {.panel-tabset}
   
   ### R
   Access the corresponding task `?mlr3::mlr_tasks_spam`.
   
   ### Python
   Read [spam.csv](https://github.com/slds-lmu/lecture_i2ml/blob/master/exercises/data/spam.csv).
   
   :::
   </details>

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details> 
<summary>**Solution**</summary>
The `spam` data is a binary classification task where the aim is to classify an e-mail as spam or non-spam:

::: {.panel-tabset}
### R
{{< embed sol_forests_R.ipynb#2-a echo=true >}}
### Python
Load data
{{< embed sol_forests_py.ipynb#2-a-1 echo=true >}}

Inspect
{{< embed sol_forests_py.ipynb#2-a-2 echo=true >}}
:::
</details> 
:::
:::

b) ::: {.callout-tip icon=false title="Only for lecture group B"}
   Use a decision tree to predict `spam`. Re-fit the tree using two random subsets of the data (each comprising 60% of observations). How stable are the trees?
   :::

   <details> 
   <summary>*Hint*</summary>
   ::: {.panel-tabset}
   
   ### R
   Use `rpart.plot()` from the package `rpart.plot` to visualize the trees.
   
   ### Python
   Use `from sklearn.tree import plot_tree` to visualize the trees.
   :::
   </details>


::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details> 
<summary>**Solution**</summary>

::: {.panel-tabset}
### R
{{< embed sol_forests_R.ipynb#2-b echo=true >}}
### Python
Full dataset
{{< embed sol_forests_py.ipynb#2-b-1 echo=true >}}

Data subsets
{{< embed sol_forests_py.ipynb#2-b-2 echo=true >}} 
:::

Observation: trees trained on different samples differ considerably in their structure, regarding split variables as well as thresholds (recall, though, that the split candidates are a further source of randomness).

</details> 
:::
:::

c) Forests come with a built-in estimate of their generalization ability via the out-of-bag (OOB) error.

::: {.indent-1-subitem}
i. Show that the probability for an observation to be OOB in an arbitrary bootstrap sample converges to $\tfrac{1}{e}$.
:::

::: {.content-visible when-profile="solution"}
::: {.indent-2-sol}
<details> 
<summary>**Solution**</summary>

This requires a little trick.
- First, think about the probability of an observation to be OOB in a tree. Imagine the tree's bootstrap sample has $n$ free spots, to be filled from the training observations.
- In each place, the probability of being drawn for the observation is $\tfrac{1}{n}$ (all observations are equally likely to be selected). Conversely, the probability of *not* being drawn is $1 - \tfrac{1}{n}$.
- We draw with replacement, meaning the events of filling a place in the bootstrap sample are all independent. The probability of not being drawn at all for any of the free spots -- i.e., not ending up in the tree's bootstrap sample and thus being OOB -- is thus $\left(1- \tfrac{1}{n}\right)^n$.
- You can imagine that this probability is lower if we only have a few observations. It will converge to a fixed value for larger datasets.
- Since we're interested in a general statement, we look for this stable value, taking $n$ to the limit: $\lim_{n \to \infty} \left( 1 - \tfrac{1}{n} \right)^n$.
- Now comes the trick: If you're well-versed in analysis you might recognize this expression as a way to characterize the exponential function.
- For an arbitrary input $x$, we have $e^x = \lim_{n \to \infty} \left( 1 + \tfrac{x}{n} \right)^n$.
- We see that our above probability is equivalent to the exponential function at input value -1, resulting in
$$
\lim_{n \to \infty} \left( 1 - \tfrac{1}{n} \right)^n = e^{-1} = 
\tfrac{1}{e} \approx 0.37.
$$

</details> 
:::
:::

::: {.indent-1-subitem}
ii. Use the random forest learner (R: `classif.ranger`, Python: `RandomForestClassifier()` to fit the model and state the out-of-bag (OOB) error.
:::

::: {.content-visible when-profile="solution"}
::: {.indent-2-sol}
<details> 
<summary>**Solution**</summary>

The OOB error can be computed by:

::: {.panel-tabset}
### R
{{< embed sol_forests_R.ipynb#2-c echo=true >}}
### Python
{{< embed sol_forests_py.ipynb#2-c echo=true >}}
:::

</details> 
:::
:::

d) You are interested in which variables have the greatest influence on the prediction quality. Explain how to determine this in a permutation-based approach and compute the importance scorses for the \texttt{spam} data.

   <details> 
   <summary>*Hint*</summary>
   ::: {.panel-tabset}
 
   ### R
   Use an adequate variable importance filter as described [here](https://mlr3filters.mlr-org.com/#variable-importance-filters).
 
   ### Python
   Choose an adequate importance measure as described [here](https://scikit-learn.org/stable/auto_examples/ensemble/plot_forest_importances.html).
   :::
   </details>

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details> 
<summary>**Solution**</summary>
Variable importance in general measures the contributions of features to a model. One way of computing the variable importance of the $j$-th variable is based on permuting it for the OOB observations and calculating the mean increase in OOB error this permutation entails.

In order to determine the with the biggest influence on prediction quality, we can choose the $k$ variables with the highest importance score, e.g., for $k = 5$:

::: {.panel-tabset}
### R
{{< embed sol_forests_R.ipynb#2-d echo=true >}}
### Python
Numerical scores
{{< embed sol_forests_py.ipynb#2-d-1 echo=true >}} 

Visual representation
{{< embed sol_forests_py.ipynb#2-d-2 echo=true >}}
:::

</details> 
:::
:::


## Exercise 3: Proximities

::: {.callout-note title="Learning goals" icon=false}
1) Be able to make predictions from code output for RF
2) Compute proximities
:::

You solve the `wine` task, predicting the \texttt{type} of a wine -- with 3 classes -- from a number of covariates. After training, you wish to determine how similar your observations are in terms of proximities.

The model information was created with 
`ranger::treeInfo()`, which assigns observations with values 
larger than `splitval` to the right child node in each split.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align="center", fig.height=5, fig.width=5}

library(data.table)
library(mlr3verse)

set.seed(123L)
task <- tsk("wine")
task$select(
  c("alcalinity", "alcohol", "flavanoids", "hue", "malic", "phenols")
)
n_trees <- 3L
learner <- lrn("classif.ranger", num.trees = n_trees, max.depth = 2L)
learner$train(task)

x_sample <- task$data()[sample(seq_len(task$nrow), 3L)]
x_sample[, type := NULL][, id := .I]
setcolorder(x_sample, "id")

  knitr::kable(
    x_sample, 
    escape = FALSE,
    col.names = c(
      "observation", 
      "alcalinity", 
      "alcohol", 
      "flavanoids", 
      "hue", 
      "malic",
      "phenols"
    ),
    # align = c("|r", "r", "r", "r", "r", "r", "r|"),
    digits = 2
    )
```
    
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align="center", fig.height=5, fig.width=5}
library(rpart.plot)

rf <- invisible(
  lapply(
    seq_len(n_trees),
    function(i) ranger::treeInfo(learner$model, tree = i)
  )
)

print("Tree 1:")
knitr::kable(
	rf[[1]], 
	escape = FALSE,
	digits = 2L 
	# align = c("|r", rep("r", 6L), "r|")
)

print("Tree 2:")
knitr::kable(
	rf[[2]], 
	escape = FALSE,
	digits = 2L 
	# align = c("|r", rep("r", 6L), "r|")
)

print("Tree 3:")
knitr::kable(
	rf[[3]], 
	escape = FALSE,
	digits = 2L 
	# align = c("|r", rep("r", 6L), "r|")
)


```

For the following subset of the training data and the random forest model given above,

a) find the terminal node of each tree the observations are placed in,

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details> 
<summary>**Solution**</summary>
Using the `treeInfo()` output, we can follow the path of each sample through each tree.
The following table prints for each observation (rows) their terminal nodes as assigned by trees 1-3. For example, consider observation 1 in tree 1 (first cell): the observation has `phenols` $> 1.94$, putting it in node 2 (`rightChild` of node 0), from there in node 6 (because it has `alcohol` $> 13.04$).

{{< embed sol_forests_py.ipynb#3-a echo=false >}}

</details> 
:::
:::

b) compute the observations' pairwise proximities, and

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details> 
<summary>**Solution**</summary>
For the proximities, we consider each pair of observations and compute the relative frequency of trees assigning them to the same terminal node. 

- Observations 1 and 2: only tree 1 assigns them to the same node, so the proximity is $\tfrac{1}{3}$.

- Observations 1 and 3: all trees assign them to the same node, so the proximity is 1.

- Observations 2 and 3: only tree 1 assigns them to the same node, so 
the proximity is $\tfrac{1}{3}$.

</details> 
:::
:::

c) construct a similarity matrix from these proximities in `R` resp. `Python`.

::: {.content-visible when-profile="solution"}
::: {.indent-1-sol}
<details> 
<summary>**Solution**</summary>
We can put this information into a similarity matrix (as such matrices become large quite quickly for more data, it is common to store only the lower diagonal -- the rest is non-informative/redundant):

::: {.panel-tabset}
### R
{{< embed sol_forests_R.ipynb#3-c-1 echo=true >}}
### Python
{{< embed sol_forests_py.ipynb#3-c-1 echo=true >}}
:::

</details> 
:::
:::
